### 2020-02-21 Writing/disparity_review.tex
added morphospace definition

@@

@@

\noindent Once suitable trait data have been collected, the design of the disparity analysis itself needs to be considered.
Study design encompasses several key aspects including (a) the difficulty of dealing with multidimensional data; (b) the variety of ordination (dimension reduction) techniques available and their limitations; (c) the metrics used to summarise the relative disparity of groups; (d) the methods used for hypothesis testing within the disparity analysis framework; and (e) ancestral state estimations in disparity-through-time analyses.
% @@@ Check the numbering here
We consider these aspects in turn below.

\subsection{a) To ordinate or not to ordinate - that is the (multidimensional) question}

@@

% -But mainly ordination is good for visualising.
One of the reason why ordinations techniques are common in disparity analysis is that they make it easier for researchers to comprehend patterns in two or three spatial dimensions at the time which can be more intuitively captured than through disparity indices (see section below). %@@@ LINK
Additionally, after ordinating the data, it is possible to focus on just a subset of axes of the morphospace (i.e. selecting only those axes that describe the majority of the variation in the dataset - e.g. 95\%).
In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions \citep{Legendre2012-va,dryden2016statistical}.
In practice, this can allow to interpret only the major axis of a highly dimensional dataset as major gradients of biological variation \citep[e.g. the elongation and flattening of birds beaks;][]{Cooney2017-ly}.

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/figure_space.pdf}
\caption{
    Morphospaces: different mathematical representations of a morphospace.
    % SG: The morphospace is the original trait space within which lie the specimens. Then one performs an ordination to visualize the space. So I would distinguish the morphospace from the ordination space. See also Mitteroecker and Huttegger (2009)
    A trait matrix can be transformed into a distance matrix \citep[e.g in][]{Close2015-qi} or an ordinated matrix \citep[e.g. in][]{Brusatte2008-vx}
    In each case, all three matrices represent the morphospace for the observations at hand.
    Visualisation: different ways to represent the morphospace in 2D.
    Visualisations can use either trait plots (directly from the trait matrix); ``flattening'' ordination of the data (e.g. using multidimensional scaling forcing all the variance to be contained in two dimensions)
    % SG: No, the flattening will not force all the variance to be contained in 2D. First because NMDS is non metric and variance is therefore not preserved as such. Second, because there is a stress associated with the flattening that correlates with the degree of distortion imposed on the data to enforce the 2D representation. Some information (variance) is thus necessarily lost.
    ; or ordination axis plots (directly from the ordinated matrix).
    % ES: I feel like somewhere we should say that in order for the spaces to be Euclidean, the axes must be on same scale. My little bugbear.
    % SG response: I think there are more conditions to be Euclidean (inner product defining the  Euclidean norm..). But surely if axes have different scales and units, it will be difficult to define a distance function and the space won't be metric.
    % @@@ Change references
}
\label{Fig:morphospace}
\end{figure}

>>>>

%@@@ add links to figures

@@

\subsection{From data to morphospace}

The concept of what the morphospace \textit{is} can be different between authors and studies but generally used in the common (i.e. non-mathematical) sense of a space containing all possible morphologies.
For the purpose of this review, we use a generalised version of the definition of \cite{mitteroecker2009concept} where the morphospace is a \textit{p}-dimensional space (or Q-space) containing \textit{n} points \citep{mitteroecker2009concept}.
In other words, mathematically, the morphospace is the mathematical object that contains the trait values (\textit{p}, the axes of the space) for all the observations (\textit{n}, the specimens); or, biologically, the morphospace contains all the observed combination of traits.
We generalise this definition to any transformation of this space that retains the biological definition, i.e. the morphospace is the mathematical object that contains the observed combinations of traits or their transformation.
Note that here the morphospace does not need to be Euclidean (although this is often desired or assumed - see discussion below). %link
If the morphospace is transformed mathematically (i.e. the dataset is transformed using logs, ordinations, distance matrices, etc...) it remains the morphospace as long as all the observation and traits remain in the datatset.

@@

\noindent Once suitable trait data have been collected, the design of the disparity analysis itself needs to be considered.
Study design encompasses several key aspects including (a) the difficulty of dealing with multidimensional data; (b) the indices used to summarise the relative disparity of groups; (c) the methods used for hypothesis testing within the disparity analysis framework; and (d) ancestral state estimations in disparity-through-time analyses.
We consider these aspects in turn below.

\subsection{(a) To ordinate or not to ordinate - that is the (multidimensional) question}

@@

One of the reason why ordinations techniques are common in disparity analysis is that they make it easier for researchers to comprehend patterns in two or three spatial dimensions at the time which can be more intuitively captured than through disparity indices (see section below). %@@@ LINK
Additionally, after ordinating the data, it is possible to focus on just a subset of axes of the morphospace (i.e. selecting only those axes that describe the majority of the variation in the dataset - e.g. 95\%).
In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions \citep{Legendre2012-va,dryden2016statistical}.
In practice, this can allow to interpret only the major axis of a highly dimensional dataset as major gradients of biological variation \citep[e.g. the elongation and flattening of birds beaks;][]{Cooney2017-ly}.

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/figure_space.pdf}
\caption{
    Morphospaces: different mathematical representations of a morphospace.
    A trait matrix can be transformed into a distance matrix \citep[e.g in][]{Close2015-qi} or an ordinated matrix \citep[e.g. in][]{Brusatte2008-vx}
    In each case, all three matrices represent the morphospace for the observations at hand.
    Visualisation: different ways to represent the morphospace in 2D.
    Visualisations can use either trait plots (directly from the trait matrix); ``flattening'' ordination of the data (e.g. using multidimensional scaling forcing all the variance to be contained in two dimensions)
    % SG: No, the flattening will not force all the variance to be contained in 2D. First because NMDS is non metric and variance is therefore not preserved as such. Second, because there is a stress associated with the flattening that correlates with the degree of distortion imposed on the data to enforce the 2D representation. Some information (variance) is thus necessarily lost.
    ; or ordination axis plots (directly from the ordinated matrix).
    % ES: I feel like somewhere we should say that in order for the spaces to be Euclidean, the axes must be on same scale. My little bugbear.
    % SG response: I think there are more conditions to be Euclidean (inner product defining the  Euclidean norm..). But surely if axes have different scales and units, it will be difficult to define a distance function and the space won't be metric.
    % @@@ Change references
}
\label{Fig:morphospace}
\end{figure}
